A matrix-free approach for finite-strain hyperelastic problems using geometric multigrid

### 2019-02-01 doc/large_strain.tex
Changes to results, conclusions, acknowledgements

%
% Below are are implementational details that might affect the comparative result
%
The results of the matrix-free approach with a geometric multigrid preconditioner are compared to the matrix-based approach in conjunction with the algebraic multigrid (AMG) preconditioner using the package ML \cite{Gee2006a} of the Trilinos \cite{Heroux2005} library.
The aggregation threshold for the AMG preconditioner is taken as $10^{-4}$.
{\color{red}
The geometric multigrid algorithm uses a V-cycle with two pre- and post-smoothing steps using Chebyshev polynomials \cite{Varga2009} of fourth degree, provided
by the \texttt{deal.II} library via the \textit{PreconditionChebyshev} class.
The largest eigenvalue $\lambda_{\rm{max}}$ of the level operators is estimated from $30$ iterations of the CG solver and the smoother
is set to target the range $[0.6\lambda_{\rm{max}}, 1.2\lambda_{\rm{max}}]$.
On the coarsest level we use {\color{red}the Chebyshev preconditioner as a solver \cite{Varga2009}} to reduce the residual by three orders of magnitude.
Since this smoother only uses diagonal elements of the operator it is readily compatible with the matrix-free approach while most default smoothers rely on an explicit matrix form.}

@@

Next we evaluate the performance of the proposed geometric multigrid preconditioner.
{\color{red}
Our main goal is to examine whether or not the adopted level operators lead to an efficient preconditioner from the linear algebra perspective. To that end}
we consider {\color{red}the average number of CG iterations throughout the entire simulation (i.e. each Newton-Raphson iteration and each loading step).}
Figures \ref{fig:benchmark_miehe_Emmy_cg2}, \ref{fig:benchmark_miehe_Emmy_cg3}, \ref{fig:benchmark_miehe_IWR_cg2} and \ref{fig:benchmark_miehe_IWR_cg3} show a
{\color{red} noticeable} increase in the average number of CG iterations for different polynomial degrees
{\color{red} for the GMG preconditioner.
Although not reported in this section, we identified that
the coarse level solver setup will greatly affect the performance of the preconditioner.
In particular, we can improve the number of iterations by choosing a more accurate coarse level solver,
however this leads to an overall more expensive linear solver.}
The black-box AMG preconditioner
{\color{red} demonstrate the same behavior of the number of CG iterations, however it}
requires many more iterations for convergence.
{\color{red}
Consequently, we can conclude that, although the adopted level operators are not built using the triple product $\gz A^{l+1}=\gz I^{l+1}_{l} \gz A^l \gz I^l_{l+1}$, they still result in a multigrid preconditioner which requires much fewer CG iterations as compared to the algebraic multigrid for the here considered finite strain elasticity problem and that the considered smoother is suitable.
}

{\color{red}
Additionally, we compare solution times of the preconditioned iterative solvers.
Our goal is to see whether the matrix-free approach with GMG preconditioner can be faster than the matrix-based approach with an AMG preconditioner.}
Figures \ref{fig:benchmark_miehe_Emmy_sol2}, \ref{fig:benchmark_miehe_Emmy_sol3}, \ref{fig:benchmark_miehe_IWR_sol2} and \ref{fig:benchmark_miehe_IWR_sol3} confirm that the efficiency of the proposed GMG preconditioner translates into faster solution times for {\color{red}all} polynomial degrees.
{\color{red}Importantly, the wall time per DoF for the matrix-free solver with GMG preconditioner stays almost constant both for 2D and 3D problems.
\changeJP{In comparison}, the wall time of the matrix-based iterative solver grows rapidly with increasing polynomial degree.
Clearly this is related both to the performance of the preconditioner from the linear algebra perspective as well as the matrix-vector products, studied in the previous section.
Based on these results, we conclude that the matrix-free approach with the adopted here geometric multigrid preconditioner can deliver a very competitive solution strategy for engendering problems
with finite strain material models.
Extension of this study from the node level to the cluster level is beyond the scope of this work,
however we hypothesize that the matrix-free and multigrid frameworks provided by \texttt{deal.II} library would provide a competitive solution for that case as well.}

@@

In this contribution, we proposed and numerically investigated {\color{red}three different} matrix-free implementations of tangent operators for finite-strain elasticity with heterogeneous materials.
{\color{red}
To the best of our knowledge, this is the first work that applies matrix-free sum factorization techniques to partial differential equations arising in this case.}
{\color{red}Among the three algorithms,}
the implementation that caches the material part of the fourth-order spatial tangent stiffness tensor together with the second-order Kirchhoff stress was shown to be faster than the matrix-based approach for quadratic elements in 3D and cubic {\color{red}elements in 2D}.
{\color{red}
Given that in this case we did not utilize the specific form of the material part of the stiffness tensor, we conclude}
that the matrix-free implementation of the tangent operator can be applied to any constitutive model which operates with the material part of the fourth-order spatial tangent stiffness tensor on the quadrature point level.

@@

D.~Davydov, D.~Arndt and J-P.~Pelteret are grateful to {\color{red}Martin Kronbichler (TU Munich),} Jed Brown (CU Boulder) and Veselin Dobrev (Lawrence Livermore National Laboratory) for a fruitful discussion on matrix-free operator evaluation approaches.

@@

>>>>

%
% Below are are implementational details that might affect the comparative result
%
\changeJP{We conduct two performance studies using the described discretization, the first being for only matrix-vector multiplication, and the second for the preconditioned iterative linear solver.}
The results of the matrix-free approach with a geometric multigrid preconditioner are compared to the matrix-based approach in conjunction with the algebraic multigrid (AMG) preconditioner using the package ML \cite{Gee2006a} of the Trilinos \cite{Heroux2005} library.
The aggregation threshold for the AMG preconditioner is taken as $10^{-4}$.
{\color{red}
The geometric multigrid algorithm uses a V-cycle with two pre- and post-smoothing steps using Chebyshev polynomials \cite{Varga2009} of fourth degree, provided
by the \texttt{deal.II} library via the \textit{PreconditionChebyshev} class.
The largest eigenvalue $\lambda_{\rm{max}}$ of the level operators is estimated from $30$ iterations of the CG solver and the smoother
is set to target the range $[0.6\lambda_{\rm{max}}, 1.2\lambda_{\rm{max}}]$.
On the coarsest level we use {\color{red}the Chebyshev preconditioner as a solver \cite{Varga2009}} to reduce the residual by three orders of magnitude.
Since this smoother only uses diagonal elements of the operator it is readily compatible with the matrix-free approach while most default smoothers rely on an explicit matrix form.}

@@

Next we evaluate the performance of the proposed geometric multigrid preconditioner.
{\color{red}
Our main goal is to examine whether or not the adopted level operators lead to an efficient preconditioner from the linear algebra perspective. To that end}
we consider {\color{red}the average number of CG iterations throughout the entire simulation (i.e. each Newton-Raphson iteration and each loading step).}
Figures \ref{fig:benchmark_miehe_Emmy_cg2}, \ref{fig:benchmark_miehe_Emmy_cg3}, \ref{fig:benchmark_miehe_IWR_cg2} and \ref{fig:benchmark_miehe_IWR_cg3} show a
{\color{red} noticeable} increase in the average number of CG iterations for different polynomial degrees
{\color{red} for the GMG preconditioner.
Although not reported in this section, we identified that
the coarse level solver setup will greatly affect the performance of the preconditioner.
In particular, we can improve the number of iterations by choosing a more accurate coarse level solver,
however this leads to an overall more expensive linear solver.}
\changeJP{However, the settings used in this study are optimized for the considered problem in terms of the wall-time spent in the linear solver.}
The black-box AMG preconditioner
{\color{red} demonstrates the same \changeJP{trend in terms of} the number of CG iterations, however it}
requires many more iterations for convergence.
\changeJP{
Consequently, we can conclude that for the here considered finite strain elasticity problem (i) the selected smoother is suitable, and that (ii) although the adopted level operators are not built using the triple product $\gz A^{l+1}=\gz I^{l+1}_{l} \gz A^l \gz I^l_{l+1}$, they still result in a multigrid preconditioner which requires much fewer CG iterations as compared to the algebraic multigrid.
}

{\color{red}
Additionally, we compare solution times of the preconditioned iterative solvers.
Our goal is to \changeJP{determine} whether the matrix-free approach with GMG preconditioner can be faster than the matrix-based approach with an AMG preconditioner.
}
\changeJP{We note that although the well-established implementation of the AMG preconditioner \cite{Gee2006a} is the best available choice for the matrix-based implementation of the boundary value problem, the current implementation of the matrix-free GMG preconditioner is not yet optimized and can be improved further.}
Figures \ref{fig:benchmark_miehe_Emmy_sol2}, \ref{fig:benchmark_miehe_Emmy_sol3}, \ref{fig:benchmark_miehe_IWR_sol2} and \ref{fig:benchmark_miehe_IWR_sol3} confirm that the efficiency of the proposed GMG preconditioner translates into faster solution times for {\color{red}all} polynomial degrees.
{\color{red}Importantly, the wall time per DoF for the matrix-free solver with GMG preconditioner stays almost constant both for 2D and 3D problems.
\changeJP{In comparison}, the wall time of the matrix-based iterative solver grows rapidly with increasing polynomial degree.
Clearly this is related both to the performance of the preconditioner from the linear algebra perspective as well as the matrix-vector products, studied in the previous section.
Based on these results, we conclude that the matrix-free approach with the adopted here geometric multigrid preconditioner can deliver a very competitive solution strategy for \changeJP{engineering problems
with compressible hyperelastic} finite strain material models.
Extension of this study from the node level to the cluster level is beyond the scope of this work,
however we hypothesize that the matrix-free and multigrid frameworks provided by \texttt{deal.II} library would provide a competitive solution for that case as well.}

@@

In this contribution, we proposed and numerically investigated {\color{red}three different} matrix-free implementations of tangent operators for finite-strain elasticity with heterogeneous materials.
{\color{red}
To the best of our knowledge, this is the first work that applies matrix-free sum factorization techniques to partial differential equations arising in this case.}
{\color{red}Among the three \changeJP{examined} algorithms,}
the implementation that caches the material part of the fourth-order spatial tangent stiffness tensor together with the second-order Kirchhoff stress was shown to be faster than the matrix-based approach for quadratic elements in 3D and cubic {\color{red}elements in 2D}.
{\color{red}
Given that in this case we did not utilize the specific form of the material part of the stiffness tensor, we conclude}
that the matrix-free implementation of the tangent operator can be applied to any constitutive model which operates with the material part of the fourth-order spatial tangent stiffness tensor on the quadrature point level.

@@

D.~Davydov, D.~Arndt and J-P.~Pelteret are grateful to {\color{red}Martin Kronbichler (TU Munich),} Jed Brown (CU Boulder) and Veselin Dobrev (Lawrence Livermore National Laboratory) for \changeJP{fruitful discussions} on matrix-free operator evaluation approaches.

@@

\todo{Add back of the envelope calculations by Veselin Dobrev? See {https://github.com/CEED/Forum/issues/1}.}
